### 2014-09-19 text/Chapter02.tex
Introduction in Chapter 4
Categorical semantics of flat calculus (Chapter 4)
Cleanup files, fix some invalid references

@@

We start by revisiting practical applications and existing language features that are related to 
coeffects (Section~\ref{sec:path-apps}), then we look at coeffects as the dual of effect systems
(Section~\ref{sec:path-eff}) and extend the duality to category theory, looking at the categorical 
dual of monads known as \emph{comonads} (Section~\ref{sec:path-sem}). Finally we look at logically 
inspired type systems that are closely related to our structural coeffects 
(Section~\ref{sec:path-logic}).

@@

As demonstrated by examples in the Chapter~\ref{ch:introduction}, this means that the resource
can be captured when a function is declared (e.g.~when it is constructed on the server-side
where database access is available), or when a function is called (e.g.~when a function created
on server-side requires access to current time-zone, it can use the resource available on the
client-side).

@@

\subsection{Effectful languages and meta-languages}

@@

\begin{definition}
A \emph{comonad} over a category $\catc$ is a triple $(C, \ident{counit}, \ident{cobind})$ where:
\begin{compactitem}
\item $C$ is a mapping on objects (types) $C : \catc \rightarrow \catc$
\item $\ident{counit}$ is a mapping $\ctyp{}{\alpha} \rightarrow \alpha$ 
\item $\ident{cobind}$ is a mapping $(\ctyp{}{\alpha} \rightarrow \beta) 
  \rightarrow (\ctyp{}{\alpha} \rightarrow \ctyp{}{\beta})$
\end{compactitem}
such that, for all $f:\alpha \rightarrow \mtyp{}{\beta}, g:\beta \rightarrow \mtyp{}{\gamma}$:
\begin{align}
\tag{\emph{left identity}}
  \ident{cobind}~\ident{counit} &= \idf{}
  \\
\tag{\emph{right identity}}
  \ident{counit} \circ \ident{cobind}~f &= f
  \\
\tag{\emph{associativity}}
  \ident{cobind}~(\ident{cobind}~g \circ f) &= (\ident{cobind}~f) \circ (\ident{cobind}~g)
\end{align}
\end{definition}

>>>>

% "Language is a labyrinth of paths. You approach from one side and know your way about; you 
% approach the same place from another side and no longer know your way about.”

@@

We start by looking at coeffects as the dual of effect systems (Section~\ref{sec:path-eff}). Next,
we extend the duality to category theory, looking at the categorical dual of monads known as 
\emph{comonads} (Section~\ref{sec:path-sem}). Finally we look at logically inspired type systems 
that are closely related to our structural coeffects (Section~\ref{sec:path-logic}).

@@

As demonstrated by examples in the Chapter~\ref{ch:intro}, this means that the resource
can be captured when a function is declared (e.g.~when it is constructed on the server-side
where database access is available), or when a function is called (e.g.~when a function created
on server-side requires access to current time-zone, it can use the resource available on the
client-side).

@@

\subsection{Effectful languages and meta-languages}
\label{sec:path-sem-langs}

@@

\todo{Maybe remove this from here, we repeate it in section 4}

\begin{definition}
A \emph{comonad} over a category $\catc$ is a triple $(C, \ident{counit}, \ident{cobind})$ where:
\begin{compactitem}
\item $C$ is a mapping on objects (types) $C : \catc \rightarrow \catc$
\item $\ident{counit}$ is a mapping $\ctyp{}{\alpha} \rightarrow \alpha$ 
\item $\ident{cobind}$ is a mapping $(\ctyp{}{\alpha} \rightarrow \beta) 
  \rightarrow (\ctyp{}{\alpha} \rightarrow \ctyp{}{\beta})$
\end{compactitem}
such that, for all $f:\ctyp{}{\alpha} \rightarrow \beta$ and $g:\ctyp{}{\beta} \rightarrow \gamma$:
\begin{align}
\tag{\emph{left identity}}
  \ident{cobind}~\ident{counit} &= \idf{}
  \\
\tag{\emph{right identity}}
  \ident{counit} \circ \ident{cobind}~f &= f
  \\
\tag{\emph{associativity}}
  \ident{cobind}~(g \circ \ident{cobind}~f) &= (\ident{cobind}~g) \circ (\ident{cobind}~f)
\end{align}
\end{definition}

\todo{The following text is duplicated in chapter 4}

### 2014-09-19 text/Chapter03.tex
Introduction in Chapter 4
Flat type system (Chapter 4)
Calculus and semantics work (Chapter 5)
Examples of structural comonads (Chapter 5)
When is coeffect a monad (Chapter 4)
Cleanup files, fix some invalid references

@@

\begin{figure}
\[
\tyrule{abs-pure}
  { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2}
  { \Gamma \vdash \lambda x.e : \tau_1 \rightarrow \tau_2 1}
\;\;
\tyrule{abs-eff}
  { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2 \,\&\, \aclrd{\sigma} }
  { \Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{\sigma}} \tau_2 \,\&\, \aclrd{\emptyset} }
\]
\caption{Lambda abstraction for pure and effectful computations}
\label{fig:applications-abs}
\end{figure}

@@

\section{Flat coeffect systems}

@@

\subsection{Implicit parameters and type classes}

@@

As we discuss later in Section~3.?, the reader monad can be extended to model rebinding.
However, later examples in this chapter, such as liveness in Section~\ref{sec:applications-flat-live}
show that other context-aware computations cannot be captured by \emph{any} monad.

@@

In addition to the work discussed already, ML5 \cite{app-distributed-ml5} is another important work 
that looks at tracking of execution environments. It uses modalities of modal S4 to represent the 
environment -- this approach is similar to coeffects, both from the practical perspective, but also 
through deeper theoretical links. We return to this topic in Chapter~\ref{ch:meta}.

@@

\paragraph{Forward vs. backward \& may vs. must.}
Static analyses can be classified as either \emph{forward} or \emph{backward} (depending on how they 
propagate information) and as either \emph{must} or \emph{may} (depending on what properties they
guarantee). Liveness is a \emph{backward} analysis -- the requirements are propagated from variables 
to their declarations. The distinction between \emph{must} and \emph{may} is apparent when we look 
at an example with conditionals:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{defaultArg}~= \lambda \ident{cond} \rightarrow \lambda \ident{input} \rightarrow\\
\quad\kvd{if}~\ident{cond}~\kvd{then}~42~\kvd{else}~\ident{input}
\end{array}
\end{equation*}
%
Liveness analysis is a \emph{may} analysis meaning that it marks variable as live when it
\emph{may} be used and as dead if it is \emph{definitely} not used. This means that the variable
\ident{input} is \emph{live} in the example above. A \emph{must} analysis would mark the variable
only if it was used in both of the branches (this is sometimes called \emph{neededness}).

The distinction between \emph{may} and \emph{must} analyses demonstrates the importance of 
interaction between contextual properties and certain language constructs such as conditionals.
We discuss this in Section~\ref{sec:flat-conditionals}

@@

\section{Structural coeffect systems}

@@

\item We choose to attach a vector of annotations to a vector of variables, rather than attaching
  individual annotations to individual variables. This lets us unify and combine flat and 
  structural systems as discussed in Chapter~\ref{ch:unified}, but the alternative is briefly
  explored in Chapter~\ref{ch:meta}.
\end{itemize}

@@

\subsection{Bounded variable use}

@@

Using the formulation of bounded logic (and omitting the terms), the weakening and
contraction rules are written as follows:
%
\[
\tyrule{weak}
  {\Gamma \vdash B}
  {\Gamma, !_0 A \vdash B}
\quad
\tyrule{contr}
  {\Gamma, !_n A, !_m A \vdash B}
  {\Gamma, !_{n+m} A \vdash B}
\]
%
The system captures the same idea as the structural coeffect system presented above.
Variable access in bounded linear logic is simply an operation that produces a value
$!_n A$ and so the system further introduces \emph{dereliction} rule which lets us 
treat $!_1 A$ as a value $A$. We further explore difference between \emph{language
semantics} and \emph{meta-language} and also revisit the BLL example in 
Chapter~\ref{ch:meta}.

@@

\paragraph{Calculus of capabilities.}
Crary et al. \cite{app-capabilities} introduced the Calculus of Capabilities to provide 
a sound system with region-based memory management for low-level code that can be easily 
compiled to assembly language. They build on the work of Tofte and Talpin \cite{app-region-memory}
who developed an effect system (as discussed in Section~\ref{sec:path-sem-effects}) that uses
lexically scoped \emph{memory regions} to provide an efficient and controlled memory management.

>>>>

% ===================================================================================================
%	                                                               
% 	  ###    #                           #                         
% 	 #   #   #                           #                         
% 	 #      ####   # ##   #   #   ###   ####   #   #  # ##    ###  
% 	  ###    #     ##  #  #   #  #   #   #     #   #  ##  #  #   # 
% 	     #   #     #      #   #  #       #     #   #  #      ##### 
% 	 #   #   #  #  #      #  ##  #   #   #  #  #  ##  #      #     
% 	  ###     ##   #       ## #   ###     ##    ## #  #       ###  
%
% ===================================================================================================

@@

\begin{figure}
\[
\tyrule{abs-pure}
  { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2}
  { \Gamma \vdash \lambda x.e : \tau_1 \rightarrow \tau_2}
\;\;
\tyrule{abs-eff}
  { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2 \,\&\, \aclrd{\sigma} }
  { \Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{\sigma}} \tau_2 \,\&\, \aclrd{\emptyset} }
\]
\caption{Lambda abstraction for pure and effectful computations}
\label{fig:applications-abs}
\end{figure}

@@

% ===================================================================================================
%	                            
% 	 #####   ##            #    
% 	 #        #            #    
% 	 #        #     ###   ####  
% 	 ####     #        #   #    
% 	 #        #     ####   #    
% 	 #        #    #   #   #  # 
% 	 #       ###    ####    ##  
%	                            
% ===================================================================================================

\section{Flat coeffect systems}
\label{sec:applications-flat}

@@

\subsection{Implicit parameters and type classes} 
\label{sec:applications-flat-impl}

@@

As we discuss later in Section~\ref{sec:flat-related-monads}, the reader monad can be extended 
to model rebinding. However, later examples in this chapter, such as liveness in 
Section~\ref{sec:applications-flat-live} show that other context-aware computations 
cannot be captured by \emph{any} monad.

@@

In addition to the work discussed already, ML5 \cite{app-distributed-ml5} is another important work 
that looks at tracking of execution environments. It uses modalities of modal S4 to represent the 
environment -- this approach is similar to coeffects, both from the practical perspective, but also 
through deeper theoretical links. We return to this topic in Section~\ref{sec:conclusions-meta} of
the final chapter.

@@

\paragraph{Forward vs. backward \& may vs. must.}
Static analyses can be classified as either \emph{forward} or \emph{backward} (depending on how they 
propagate information) and as either \emph{must} or \emph{may} (depending on what properties they
guarantee). Liveness is a \emph{backward} analysis -- the requirements are propagated from variables 
to their declarations. The distinction between \emph{must} and \emph{may} is apparent when we look 
at an example with conditionals:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{defaultArg}~= \lambda \ident{cond} \rightarrow \lambda \ident{input} \rightarrow\\
\quad\kvd{if}~\ident{cond}~\kvd{then}~42~\kvd{else}~\ident{input}
\end{array}
\end{equation*}
%
Liveness analysis is a \emph{may} analysis meaning that it marks variable as live when it
\emph{may} be used and as dead if it is \emph{definitely} not used. This means that the variable
\ident{input} is \emph{live} in the example above. A \emph{must} analysis would mark the variable
only if it was used in both of the branches (this is sometimes called \emph{neededness}).
The distinction between \emph{may} and \emph{must} analyses demonstrates the importance of 
interaction between contextual properties and certain language constructs such as conditionals.

@@

% ==================================================================================================
%	                                                                      
% 	  ###    #                           #                           ##   
% 	 #   #   #                           #                            #   
% 	 #      ####   # ##   #   #   ###   ####   #   #  # ##    ###     #   
% 	  ###    #     ##  #  #   #  #   #   #     #   #  ##  #      #    #   
% 	     #   #     #      #   #  #       #     #   #  #       ####    #   
% 	 #   #   #  #  #      #  ##  #   #   #  #  #  ##  #      #   #    #   
% 	  ###     ##   #       ## #   ###     ##    ## #  #       ####   ###  
%	                                                                      
% ==================================================================================================

\section{Structural coeffect systems}
\label{sec:applications-structural}

@@

\item We choose to attach a vector of annotations to a vector of variables, rather than attaching
  individual annotations to individual variables. This lets us unify and combine flat and 
  structural systems as discussed in Chapter~\ref{ch:unified}, but the alternative is briefly
  explored in Section~\ref{sec:conclusions-meta}.
\end{itemize}

@@

\subsection{Bounded variable use}
\label{sec:applications-struct-bll}

@@

Using the formulation of bounded logic (and omitting the terms), the weakening and
contraction rules are written as follows:
%
\[
\tyrule{weak}
  {\Gamma \vdash B}
  {\Gamma, !_0 A \vdash B}
\quad
\tyrule{contr}
  {\Gamma, !_n A, !_m A \vdash B}
  {\Gamma, !_{n+m} A \vdash B}
\]
%
The system captures the same idea as the structural coeffect system presented above.
Variable access in bounded linear logic is simply an operation that produces a value
$!_n A$ and so the system further introduces \emph{dereliction} rule which lets us 
treat $!_1 A$ as a value $A$. We further explore difference between \emph{language
semantics} and \emph{meta-language} in Section~\ref{sec:conclusions-meta}.

@@

\paragraph{Calculus of capabilities.}
\label{sec:applications-active-ccc}

Crary et al. \cite{app-capabilities} introduced the Calculus of Capabilities to provide 
a sound system with region-based memory management for low-level code that can be easily 
compiled to assembly language. They build on the work of Tofte and Talpin \cite{app-region-memory}
who developed an effect system (as discussed in Section~\ref{sec:path-sem-effects}) that uses
lexically scoped \emph{memory regions} to provide an efficient and controlled memory management.
